---
title: "in class template, Week 5: aquatic invertebrates"
subtitle: "Key"
date: "2026-04-27"
author: "insert your name here"
format: 
  pdf:
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
execute:
  message: false
  warning: false
---

# 1. Set up

## Packages

```{r}
#| label: packages

# general use and cleaning
library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrapping axis labels
library(scales)

# reading in .xlsx files
library(readxl)

# visualizing missing data
library(naniar)

# arranging plots
library(patchwork)
```

## Data

```{r}
#| label: data

# taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv")) 

# aquatic invertebrates
aq_ins <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                    sheet = "Aquatic Insects") 

# site information
sites <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                   sheet = "sites")

# water quality information
water_quality <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                           sheet = "Water Quality",
                           na = c("", "n/a", "N/A", "over", "173+")) 
```

# 2. Cleaning

## `sites` object

Goal: create a new object from `sites` that only includes NCOS sites

Use: filter downstream data frames using new object

```{r}
#| label: ncos-site-object

# creating new object from sites
ncos_sites <- sites |> 
  # filter to only include NCOS sites sampled four times a year
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")
```

## Cleaning `taxon_list`

Goal: convert character strings in existing `taxon_list` data frame into snake case

Use: join with aquatic inverts to attach taxonomic information to species IDs

```{r}
#| label: cleaning-taxon-list

# creating new clean object from taxon_list
taxon_list_clean <- taxon_list |> 
  # converting taxon name to snake case (no spaces or capital letters,
  # only underscores)
  mutate(verbatim_name = to_any_case(verbatim_name, case = "snake"))
```

## Cleaning `water_quality`

Goal: create data frame with mean pH, mean DO (in mg/L), and mean salinity (in ppt) for each date and site

Use: join with aquatic inverts to attach water quality information to surveys

Pseudocode:

- create new object from `water_quality`
- clean column names
- only include surveys from quarterly NCOS sites
- filter to only include sites in NCOS sites
- create a new column called `date_site` to join with later data frames
- group by `date_site`
- calculate median pH, salinity (ppt), and dissolved oxygen (mg/L)
- ungroup data frame
- filter out salinity outliers

```{r}
#| label: water-quality-cleaning

# creating new clean object from water quality
water_quality_clean <- water_quality |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites in ncos_sites
  semi_join(ncos_sites, by = "site") |> 
  # create new column to join with later data frames
  unite("date_site", date, site, remove = TRUE) |> 
  # group by date_site column
  group_by(date_site) |> 
  # calculate median pH, salinity, DO
  summarize(med_ph = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # filter out salinity outliers
  filter(date_site != "2022-08-12_NVBR") 
```

## Cleaning `aq_ins`

Goal: create long format data frame in which each row contains the counted individuals in each taxon during each survey on each date with all the taxonomic information for the taxon and the water quality information for the survey

Use: visualize relationships between water quality parameters and aquatic invertebrate abundances

Pseudocode:

- create new object from `aq_ins`
- clean column names
- filter to only include sites in NCOS sites
- select columns of interest
- filter to only include "filtered beaker" samples
- convert the data frame to long format
- group by date, site, taxon
- calculate average abundance per liter (sum abundance divided by 7.5 liters)
- create a new column called `date_site` to join with water quality
- join with water quality
- join with taxon list

```{r}
#| label: cleaning-aq-ins

# creating new clean object from aquatic inverts
aq_ins_clean <- aq_ins |> 
  # clean column names
  clean_names() |> 
  # filtering join: only include sites occurring in ncos_sites
  semi_join(ncos_sites, by = c("site")) |> 
  # renaming columns
  rename(date = date_on_vial) |> 
  # select columns of interest
  select(date, sample_type, site,
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |> 
  # filter to only include "filtered beaker" samples
  filter(sample_type %in% c("FB 250", "FB250")) |> 
  # convert the data frame to long format
  pivot_longer(cols = ostracod:annelida_polychaete,
               names_to = "taxon",
               values_to = "abundance")
